---
title: "Binary logistic regression"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(aplore3)
library(tableone)
library(stargazer)
library(tidyverse)
library(ggdag)
knitr::opts_chunk$set(echo = TRUE, comment=NA)
tutorial_options(exercise.timelimit=40)
```

## Context
A cross sectional study was conducted to investigate the association between different activity levels and occurrence of myopia among school children of age between 5-9. 618 children were recruited and following variables were measured:  

variable name | description
---------------|------------
age | Age (Years)
gender | Gender (1: Male, 2: Female)
spheq | Spherical Equivalent Refraction (diopter)
al | Axial length (mm)
acd | Anterior Chamber Depth (mm)
lt | Lens Thickness (mm)
vcd | Vitreous Chamber Depth (mm)
sporthr | How many hours per week outside of school the child spent engaging in sports/outdoor activities (Hours per week)   
readhr | How many hours per week outside of school the child spent reading for pleasure (Hours per week)     
comphr | How many hours per week outside of school the child spent playing video/computer games or working on the computer (Hours per week)     
studyhr | How many hours per week outside of school the child spent reading or studying for school assignments (Hours per week)    
tvhr | How many hours per week outside of school the child spent watching television (Hours per week)    
diopterhr | Composite of near-work activities (Hours per week)     
mommy | Was the subject's mother myopic? (1: No, 2: Yes)    
dadmy | Was the subject's father myopic? (1: No, 2: Yes)  

### Research Question  
**To estimate the extend to which hours spend outside school, engagin in sports/outdoor activity, is associated with risk of myopia among children below 10 years of age**

### DAG

Assuming that there are no unmeasured confounding between myopic and other variables in the data. We can construct a DAG as below. 
We will also assume that this DAG is encode our current knowledge and is complete. 
```{r}
theme_set(theme_dag_blank())
coords = list(
        x = c(mommy=2, dadmy=2, sporthr=2, readhr=2, comphr=2, studyhr=2, tvhr=2, diopterhr=2, myopic=4,
              age=1, gender=1, spheq=3, al=3, acd=3, lt=3, vcd=3),
        y = c(mommy=-4, dadmy=-3, readhr=-2, comphr=-1, sporthr=0, studyhr=1, tvhr=2, diopterhr=3, myopic=0,
              age=-1, gender=1, spheq=-4, al=-3, acd=-2, lt=2, vcd=3)
)

dag1 <- dagify(myopic ~ age+gender+mommy+dadmy+sporthr+readhr+comphr+studyhr+tvhr+diopterhr+spheq+al+acd+lt+vcd,
               mommy ~~ sporthr ,
               dadmy ~~ sporthr,
               sporthr ~~ readhr,
               sporthr ~~ comphr,
               sporthr ~~ studyhr,
               sporthr ~~ tvhr,
               sporthr ~~ diopterhr,
               sporthr ~ age+gender,
               readhr ~ age+gender,
               comphr ~ age+gender,
               studyhr ~ age+gender,
               tvhr ~ age+gender,
               diopterhr ~ age+gender,
               spheq ~ age+gender+sporthr,
               al ~ age+gender+sporthr,
               acd ~ age+gender+sporthr,
               lt ~ age+gender+sporthr,
               vcd ~ age+gender+sporthr,
              exposure ="sporthr",
              outcome = "myopic", coords = coords)
```

```{r,fig.align='center', fig.width=10, fig.height=8}
ggdag(dag1, text_size =2.8, node=T)
```

To know which variables to adjust for in our model, we can use ```ggdag_adjustment_set()``` function.

```{r,fig.align='center', fig.width=10, fig.height=8}
ggdag_adjustment_set(dag1, text_size=2.8)
```

Because we are interested in the total effect of sporthr, we will not adjust for the descendants of sporthr. 

## Descriptive statistics

The data is available from the ```aplore3``` package. 
We can load the necessary libraries and data as follows:
```{r set2, exercise.eval=TRUE}
library(learnr)
library(aplore3)
library(tableone)
library(stargazer)
library(tidyverse)
data(myopia)
```

We look at the distribution of continuous variables

```{r bp1, exercise=TRUE}
layout(matrix(c(1:10), nrow=2, ncol=2, byrow = T))

boxplot(myopia$age, main="Boxplot of age")
hist(myopia$age)

boxplot(myopia$al, main="Boxplot of al")
hist(myopia$al)

boxplot(myopia$acd, main="Boxplot of acd")
hist(myopia$acd)

boxplot(myopia$lt, main="Boxplot of lt")
hist(myopia$lt)

boxplot(myopia$vcd, main="Boxplot of vcd")
hist(myopia$vcd)
```

**Exercise 1:** Use the same strategy to create boxplots for ```c("sporthr","readhr","comphr","studyhr","tvhr","diopterhr")```
```{r bp2, exercise=TRUE}

```

```{r bp2-solution}
layout(matrix(c(1:12), nrow=2, ncol=2, byrow = T))

boxplot(myopia$sporthr)
hist(myopia$sporthr)

boxplot(myopia$readhr)
hist(myopia$readhr)

boxplot(myopia$comphr)
hist(myopia$comphr)

boxplot(myopia$studyhr)
hist(myopia$studyhr)

boxplot(myopia$tvhr)
hist(myopia$tvhr)

boxplot(myopia$diopterhr)
hist(myopia$diopterhr)
```

+ Most of the variables have outliers 

For categorical variables we can tabulate them using the tableone package
```{r}
tab1 <- CreateTableOne(data=myopia, 
                       vars=c("myopic","gender","mommy","dadmy"), includeNA = TRUE)
print(tab1, showAllLevels=TRUE)
```

+ Only 81 (13\%) participants have myopia. 

### Bivariate statistics
### Table 1
We shall look at the distribution of all variables across the outcome.
```{r tab1, exercise=TRUE}
tab1 <- CreateTableOne(vars=c("age","gender","sporthr",
                              "readhr","comphr"),
                       strata = "myopic",
                       data=myopia,
                       includeNA=TRUE,
                       test=FALSE)

print(tab1, showAllLevels=TRUE)
```

**Excersise 2:** Create the tableone for variables ```c("studyhr","tvhr","diopterhr", "mommy","dadmy")```

```{r tab2, exercise=TRUE}

```


```{r tab2-solution}
tab2<- CreateTableOne(vars=c("studyhr","tvhr","diopterhr", "mommy","dadmy"),
                       strata = "myopic",
                       data=myopia,
                       includeNA=TRUE,
                       test=FALSE)

print(tab2, showAllLevels=TRUE)
```

+ Myopic kids had more average hours in sport/outdoor activity
+ There are slightly more Female kids among myopic group
+ Myopic kids had high proportion of participants with family history of myopia
+ There are no missing values in any of the variables

### Correlations
Because most of the continuous variables are not normally distributed we will use spearman's correlation. 

```{r cor1, exercise=TRUE}
corr1 <- corr(myopia[, c("age","spheq","al","acd","lt","vcd","sporthr")], method="spearman") %>%
round(corr1, digits = 2)
```

**Exercise 3:** Create the correlation matrix of all continuous variables in the data.
```{r cor2, exercise=TRUE}


```

```{r cor2-solution}
corr <- corr(myopia[, c("age","spheq","al","acd","lt","vcd",'sporthr',"readhr",'comphr','studyhr','tvhr','diopterhr')],
             method="spearman") %>%
round(corr, digits = 2)
```

+ vcd is highly correlated with al
+ Diopterhr variable seems to have moderate level correlations with some other activity variables

## logistic regressions

### Univariate binary logistic regressions
We will regress each independant variable on the outcome 
```{r un1, exercise=TRUE}
uniregs <- list()
uniregs[[1]] <- glm(myopic ~ age, data=myopia, family=binomial(link="logit"))
uniregs[[2]] <- glm(myopic ~ gender, data=myopia, family=binomial(link="logit"))
uniregs[[3]] <- glm(myopic ~ sporthr, data=myopia, family=binomial(link="logit"))
uniregs[[4]] <- glm(myopic ~ readhr, data=myopia, family=binomial(link="logit"))
uniregs[[5]] <- glm(myopic ~ studyhr, data=myopia, family=binomial(link="logit"))


stargazer(uniregs[1:5], type="text", report="vcs", ci=TRUE, omit.table.layout = "n", digits=2)
```

**Exercise 4:** Follow the same startegy as above and add the univariate regression outputs for rest of the potential confounders to the same object ```uniregs``` and print the coefficients.

```{r uni2, exercise=TRUE}

```


```{r uni2-solution}
uniregs[[6]] <- glm(myopic ~ tvhr, data=myopia, family=binomial(link="logit"))
uniregs[[7]] <- glm(myopic ~ diopterhr, data=myopia, family=binomial(link="logit"))
uniregs[[8]] <- glm(myopic ~ mommy, data=myopia, family=binomial(link="logit"))
uniregs[[9]] <- glm(myopic ~ dadmy, data=myopia, family=binomial(link="logit"))
uniregs[[10]] <- glm(myopic ~ comphr, data=myopia, family=binomial(link="logit"))


stargazer(uniregs[6:10], type="text", report="vcs", ci=TRUE, omit.table.layout = "n", digits=2)

```

+ age, gender(female), readhr, comphr, diopterhr, mommy, dadmy show positive association with myopic state
+ sporthr show negative association
+ interestingly, studyhr and tvhr show a negative association!

### Multivariable binary logistic regression
Now we can build our multivariable logistic regression model as follows:

```{r}
glm <- glm(myopic ~ age + gender + mommy + dadmy + 
              sporthr + readhr + comphr + studyhr + tvhr + diopterhr , 
            data=myopia, family=binomial(link="logit"))
stargazer(glm, type="text", report="vcs", ci=TRUE, omit.table.layout = "n", digits=2)
```

**The diopterhr variable do not have a coefficient. Probably due to high collinearity R took it out**  


**We need to report the effect measures in Odds ratio scale**. 
An example with only one independent variable:

```{r glm1, exercise=TRUE}
glm1 <- glm(myopic ~ sporthr, 
            data=myopia, family=binomial(link="logit"))

# Exponentiate the logodds (betas) to odds ratios

ors <- exp(coef(glm1))

# Compute their confidence intervals
cis <- exp(confint.default(glm1))

# Use the above objects in stargazer
stargazer(glm1, type="text", report="vcs", 
          coef=list(ors), 
          ci.custom = list(cis), 
          omit.table.layout = "n", 
          digits=2)
```

**Exercise 5:** Add all the potential confounders (except diopterhr) and re-fit the model and report the results in odds ratios
```{r od1, exercise=TRUE}


```

```{r od1-solution}
glm2 <- glm(myopic ~ age + gender + mommy + dadmy + 
              sporthr + readhr + comphr + studyhr + tvhr, 
            data=myopia, family=binomial(link="logit"))

# Exponentiate the logodds (betas) to odds ratios

ors2 <- exp(coef(glm2))

# Compute their confidence intervals
cis2 <- exp(confint.default(glm2))

# Use the above objects in stargazer
stargazer(glm2, type="text", report="vcs", 
          coef=list(ors2), 
          ci.custom = list(cis2), 
          omit.table.layout = "n", 
          digits=2)
```

### Interpretation 
Each hour increase in hours spend outside school, engaging in sports/outdoor activities, per week decreased the odds of myopia by a factor of 0.96 (OR = 0.06, 95\%CI = 0.93, 0.99). 
